\section{Using \textsc{Patus}}

\subsection{A Walkthrough Example}
\label{sec:wave}

In \textsc{Patus}, the user specifies the actual stencil computation.
Thus, if the tool is to be used to solve differential equations using a finite difference-based
discretization method, the discretization needs to be done prior to the implementation in \textsc{Patus}.
In the following, we show briefly how this can be done by means of a simple example.


\subsubsection{From a Model to a Stencil Specification}

Consider the classical wave equation\index{wave equation} on $\Omega=[-1,1]^3$ with Dirichlet boundary conditions and some initial condition:
\begin{align}
  \frac{\partial^2 u}{\partial t^2} - c^2 \Delta u & = 0 \qquad \text{in }\Omega, \nonumber \\
  u & = 0 \qquad \text{on }\partial\Omega, \\
  u(x,y,z)|_{t=0} & = \sin(2\pi x) \sin(2\pi y) \sin (2\pi z). \nonumber
  \label{eq:wave}
\end{align}

Using an explicit finite difference method to discretize the equation both in space and time
by means of a fourth-order discretization of the Laplacian $\Delta$ over an equidistant spatial grid with stepsize $h$
and a second-order scheme with time step $\delta t$ in time, we obtain
\begin{equation}
	\label{eq:wave-discrete}
	\frac{u^{(t+\delta t)}-2u^{(t)}+u^{(t-\delta t)}}{\delta t} - c^2 \Delta_h u^{(t)} = 0,
\end{equation}
where $\Delta_h$ is the discretized version of the Laplacian:
\begin{align}
  \label{eq:wave-discrete-laplacian}
  \Delta_h u^{(t)}(x,y,z) = -\tfrac{15}{2h^2} & u^{(t)}(x,y,z) + \\
    -\tfrac{1}{12h^2} & \left(u^{(t)}(x-2h,y,z) + u^{(t)}(x,y-2h,z) + u^{(t)}(x,y,z-2h) \right) + \nonumber \\
     \tfrac{4}{3h^2}  & \left(u^{(t)}(x-h,y,z)  + u^{(t)}(x,y-h,z)  + u^{(t)}(x,y,z-h)  \right) + \nonumber \\                     
     \tfrac{4}{3h^2}  & \left(u^{(t)}(x+h,y,z)  + u^{(t)}(x,y+h,z)  + u^{(t)}(x,y,z+h)  \right) + \nonumber \\
    -\tfrac{1}{12h^2} & \left(u^{(t)}(x+2h,y,z) + u^{(t)}(x,y+2h,z) + u^{(t)}(x,y,z+2h) \right).  \nonumber
\end{align}
Substituting Eqn. \ref{eq:wave-discrete-laplacian} into Eqn. \ref{eq:wave-discrete},
solving Eqn. \ref{eq:wave-discrete} for $u^{(t+\delta t)}$, and interpreting $u$ as a grid in space and time with mesh size $h$
and time step $\delta t$, we arrive at
\begin{align*}
  u[x,y,z;t+1] =  2 & u[x,y,z;t] - u[x,y,z;t-1] + c^2\tfrac{\delta t}{h^2} \Big( -\tfrac{15}{2} u[x,y,z;t] + \\
    -\tfrac{1}{12}& \left(u[x-2,y,z;t] + u[x,y-2,z;t] + u[x,y,z-2;t] \right) + \\
     \tfrac{4}{3}  & \left(u[x-1,y,z;t] + u[x,y-1,z;t] + u[x,y,z-1;t] \right) + \\                     
     \tfrac{4}{3}  & \left(u[x+1,y,z;t] + u[x,y+1,z;t] + u[x,y,z+1;t] \right) + \\
    -\tfrac{1}{12}& \left(u[x+2,y,z;t] + u[x,y+2,z;t] + u[x,y,z+2;t] \right) \!\!\Big).
\end{align*}
This can now be turned into a \textsc{Patus} stencil specification almost trivially:

\begin{lstlisting}[language=stencil]
stencil wave(
  float grid U(0..x_max-1, 0..y_max-1, 0..z_max-1), 
  float param fMin = -1,
  float param fDX = 2 / (x_max-3),
  float param fDT_DX_sq = 0.25)
{
  // do one timestep within the stencil kernel
  iterate while t < 1;
  
  // define the region on which the stencil is evaluated
  domainsize = (2..x_max-3, 2..y_max-3, 2..z_max-3);
  
  // define the initial condition (how the data is initialized
  // before the computation)
  // note that 0 <= x < x_max, etc.
  initial {
    U[x,y,z; -1] = sinf(2*~$\pi$~*((x-1)*fDX+fMin)) *
      sinf(2*~$\pi$~*((y-1)*fDX+fMin)) * sinf(2*~$\pi$~*((z-1)*fDX+fMin));
    U[x, y, z; -1 :
      x==0 || y==0 || z==0 || x==x_max-1 || y==y_max-1 || z==z_max-1 ] = 0;
    U[x,y,z; 0] = U[x,y,z; -1];
    U[x,y,z; 1] = 0;
  }

  // define the actual stencil computation
  operation {
    float c1 = 2 - 15/2 * fDT_DX_sq;
    float c2 = 4/3 * fDT_DX_sq;
    float c3 = -1/12 * fDT_DX_sq;
    
    U[x,y,z; t+1] = c1 * U[x,y,z; t] - U[x,y,z; t-1] +
      c2 * (
        U[x+1,y,z; t] + U[x-1,y,z; t] +
        U[x,y+1,z; t] + U[x,y-1,z; t] +
        U[x,y,z+1; t] + U[x,y,z-1; t]
      ) +
      c3 * (
        U[x+2,y,z; t] + U[x-2,y,z; t] +
        U[x,y+2,z; t] + U[x,y-2,z; t] +
        U[x,y,z+2; t] + U[x,y,z-2; t]
      );
  }
}
\end{lstlisting}

\noindent From this stencil specification, \textsc{Patus} will generate
\begin{itemize}
  \item a C source code file implementing the stencil computation and
  \item source files from which a benchmarking harness can be built.
\end{itemize}
The benchmarking executable is then used by the auto-tuner, which determines a set of
architecture-specific parameters, for which the stencil achieves the best performance.


\subsubsection{The Specification Explained}

\begin{itemize}
  \item The \textbf{stencil} specification defines a stencil ``wave,'' which operates on a grid (called ``U'')
    of size $[$0, x\_max-1$] \times [$0, y\_max-1$] \times [$0,z\_max-1$]$.
    Note that the size parameters do not have to be defined in the stencil specification.
    Instead, they will appear as arguments to the generated stencil kernel function, and, in the
    benchmarking harness, as command line arguments.
    
    The remaining arguments, ``fMin,'' ``fDX,'' and ``fDT\_DX\_sq,'' are used for initializing the grid
    and performing the stencil computation. These parameters will also appear as arguments to the generated
    C function implementing the computation.

    Optionally, parameters can be initialized with default
    values (as shown in the listing above). This affects only the benchmarking harness and will fix the
    values which are passed to the generated stencil kernel.

  \item The \textbf{iterate while} statement defines the number of timesteps to be performed within one stencil kernel call.
    In the example, one timestep per kernel invocation will be performed. (Timesteps are counted from 0.)
    The statement can be omitted. Then, the number of timesteps defaults to 1.
    In the future we will also allow (reduction-based) convergence criteria (e.g., ``iterate as long as the residual is
    larger than some $\varepsilon$'').
    
  \item The \textbf{domainsize} defines the iteration space. While the total size of the grid extends from 0 to $\ast$\_max-1,
    the stencil is only applied to the points between 2 and $\ast$\_max-3 ($\ast \in \{$x, y, z$\}$), i.e., only
    to the \emph{interior} grid points.
    
  \item The \textbf{initial} block defines how the grid is initialized, or more mathematically: it defines the initial condition
    of the (discretized) PDE. The ``sinf'' function is actually a C function (single precision sine)
    which will be called (this is the behavior
    if there is a function which is not known to \textsc{Patus}).
    Note that \textsc{Patus} also defines the $\pi$ literal with the obvious meaning.\\
    The second statement of the initialization sets all the grid points for which x, y, z are 0 or $\ast$\_max-1
    (the condition after colon) to zero. I.e., you can use the set builder notation (with any logical and comparison operators you know from C/C++)
    to select certain grid points and initialize them.\\
    \textbf{initial} blocks are not mandatory; if no \textbf{initial} is provided, \textsc{Patus} will create an initialization
    routine anyway, initializing the data with arbitrary values. (This is to ensure the correct data placement on NUMA machines.)
    
  \item The \textbf{operation}, finally, defines the actual stencil computation.
    It can contain definitions of constants (as in the listing) or temporary values, and it can also contain more than
    one stencil expression.
  
  \item Optionally, the stencil specification can also contain a \textbf{boundaries} block, in which
    special treatment of boundary regions can be specified. Essentially, within the \textbf{boundaries} block,
    special stencils are defined which are applied to boundary regions. See section \ref{sec:stencilspec}.
\end{itemize}


\subsubsection{Building the Benchmarking Harness}

Once the stencil specification is written, \textsc{Patus} can be run to transform it into C code.
Assume the stencil specification was saved in the file \texttt{examples/ stencils/wave-1.stc}.
(This very stencil specification is actually there.)
In your shell (in Linux), type
\cmd{cd examples/stencils\\
%  patus examples/stencils/wave-1.stc --outdir=out/wave-1
  \phantom{XXXX}patus wave-1.stc
}
or, in Microsoft Windows, on the command line type
\cmd{cd examples\textbackslash{}stencils\\
  \phantom{XXXX}..\textbackslash{}..\textbackslash{}bin\textbackslash{}patus.bat wave-1.stc
}
This will generate the C code implementing the Wave stencil, using the default architecture,
and putting the generated files in the
\texttt{out} directory. (You can change the output directly by adding the \texttt{--outdir=$<$your-output-dir$>$}
to the \textsc{Patus} command line.)
Change into the output directory, \cmd{cd out}
and type \cmd{make} to build the benchmarking harness, which will be used for the auto-tuning process, and can also be
used for simple simulations. When the build completes successfully, there is an executable file,
\texttt{bench} (or \texttt{bench.exe} on Windows) in the directory.

If you try to start the benchmarking executable, you will see that it expects some command line arguments:
\example{\small\texttt{%
  \$  ./bench\\
  Wrong number of parameters. Syntax:\\
  ./bench <x\_max> <y\_max> <z\_max> <cb\_x> <cb\_y> <cb\_z> <chunk> <\_unroll\_p3>
}}

The $\ast$\_max correspond to the unspecified domain size variables in the stencil specification.
All the other arguments come from the code generator, and it is the auto-tuner's task to find the best values
for them.


\subsubsection{Auto-Tuning}

The Makefile (which was also used to build the benchmarking harness) defines a ``tune'' target, which
starts the auto-tuner. ``tune'' expects the domain size variables to be specified:
\cmd{make tune x\_max=64 y\_max=64 z\_max=64}
Type the above command in the shell. The auto-tuner will run the benchmarking executable a number of times,
varying the values of the arguments to determine, and it will terminate with a message like

\example{\small
  \texttt{Optimal parameter configuration found:}\\
  \texttt{64 64 64 62 20 8 2 0}\\

  \texttt{Timing information for the optimal run:}\\
  \texttt{0.9047935972598529}\\

  \texttt{Program output of the optimal run:}\\
  \texttt{Flops / stencil call:\phantom{XX}16}\\
  \texttt{Stencil computations:\phantom{XX}1116000}\\
  \texttt{Bytes transferred:\phantom{XXXXX}15728640}\\
  \texttt{Total Flops:\phantom{XXXXXXXXXXX}17856000}\\
  \texttt{Seconds elapsed:\phantom{XXXXXXX}0.000423}\\
  \texttt{Performance:\phantom{XXXXXXXXXXX}42.193517 GFlop/s}\\
  \texttt{Bandwidth utilization:\phantom{X}37.166590 GB/s}\\
  \texttt{1352692.000000}\\
  \texttt{Validation OK.}
}

\noindent Thus, according to the auto-tuner, on the machine the benchmark was run and for x\_max=y\_max=z\_max=64,
the best parameter combination is \cmd{cb\_x=62, cb\_y=20, cb\_z=8, chunk=2, \_unroll\_p3=0.}

\textsc{Patus} has certain requirements for the grid sizes in the unit stride dimension.
If the requirements are not met when you run the executable or the auto-tuner, you may see an error message like
\cmd{Non-native, aligned SIMD type mode requires that (x\_max+2) is divisible by 4 [(x\_max+2) = 102].}
If this is the case, adjust the grid size (here, e.g., set \texttt{x\_max} to 102 instead of 100) and run again.


\subsubsection{Simple Visualization}

You can run the benchmarking executable with the arguments provided by the auto-tuner,
\cmd{./bench 64 64 64 62 20 8 2 0}
If you add the flag \texttt{-o} to the command line,
\cmd{./bench 64 64 64 62 20 8 2 0 -o}
the program will take snapshots of the data grids before and after the execution of the stencil kernel
and write them to text files. If you have gnuplot installed on your system, you can run
\cmd{make plot} which will use gnuplot to create image files from the data text files.


\subsection{Writing Your Own Stencil Specifications}
\label{sec:stencilspec}

A stencil specification has the following form:
\example{
  \texttt{stencil} \textit{\footnotesize stencil-name} \texttt{(} \textit{\footnotesize stencil-arguments} \texttt{)}
  \texttt{\{}\\
  \phantom{XXXX}\textit{\footnotesize iteration-space}\\
  \phantom{XXXX}\textit{[ {\footnotesize number-of-timesteps} ]}\\
  \phantom{XXXX}\textit{\footnotesize operation}\\
  \phantom{XXXX}\textit{[ {\footnotesize boundaries} ]}\\
  \phantom{XXXX}\textit{[ {\footnotesize initial} ]}\\
  \texttt{\}}
}

The order of the iteration space, number of timesteps, operation, boundaries, and initial specifications
is not relevant.


\subsubsection{Stencil Arguments}

Arguments to the stencil can be either \textbf{grid}s or \textbf{param}s. Two datatypes are supported:
\textbf{float} (for single precision floating point numbers) and \textbf{double} (double precision
floating point numbers). Grids are multi-dimensional arrays of data which can be read and written to.
Params are application-specific read-only data used in the computation. They can be scalars or fixed-size
arrays.

\bigskip

\noindent A \textbf{grid} can be declared as follows:
\example{
  \textit{[} \texttt{const} \textit{]}
  \textit{(} \texttt{float} $|$ \texttt{double} \textit{)}
  \texttt{grid}
  \textit{\footnotesize grid-name}
  \textit{[}\texttt{(}
  \textit{\footnotesize lbnd$_1$} \texttt{..} \textit{\footnotesize ubnd$_1$}\texttt{,}
  \textit{\footnotesize lbnd$_2$} \texttt{..} \textit{\footnotesize ubnd$_2$}
  \dots
  \texttt{)}\textit{]}\\
  \textit{[}
  \texttt{[}
  \textit{\footnotesize albnd$_1$} \textit{[}\texttt{..} \textit{\footnotesize aubnd$_1$}\textit{]} \texttt{,}
  \textit{\footnotesize albnd$_2$} \textit{[}\texttt{..} \textit{\footnotesize aubnd$_2$}\textit{]}
  \dots
  \texttt{]}
  \textit{]}
}

\noindent All the upper bounds are inclusive.

\medskip
\noindent Examples:
\begin{lstlisting}[language=stencil]
float grid A
const double grid B
float grid C(-1 .. size_x)
const double grid D(0..sx, 1..sy, -2..sz+2)[3]
float grid E(min_x..max_x, min_y..max_y)[0..1, -1..3]
\end{lstlisting}

The grid \texttt{A} is declared as a single precision grid with no explicit size declaration.
With the size declaration the size of the grid in memory can be defined, which is distinct from the
iteration domain, i.e., the specification of the domain to which the calculation is applied.
In this case, the grid's size is implied from the \textbf{domainsize} in the stencil specification:
the grid is assumed to be the smallest grid such that all the grid accesses in the stencil computation are valid.

For the grid \texttt{B} also no explicit size is declared. \textbf{const} means that the grid is read-only.

Grid \texttt{C} is declared as a one-dimensional grid indexable from -1 to size\_x.
In the actual stencil computation, not all of these points need to be written to (or read), but
no bounds checking is performed. If the computation violates the bounds, the compiled code probably will raise
segmentation faults.

\texttt{D} is declared as an array of 3 three-dimensional grids,
\texttt{E} as a two-di\-men\-sional array of two-dimensional grids.
The examples also show that the lower and upper bounds can be arbitrary arithmetic expressions containing variables.
Any variables appearing in size declarations will become arguments to the generated stencil function.
The array indices of \texttt{E} can be 0 or 1 (first component) and between (and including) -1 and 3 (second component).

\bigskip

\noindent A \texttt{param} can be declared as follows:
\example{
  \textit{(} \texttt{float} $|$ \texttt{double} \textit{)}
  \texttt{param}
  \textit{\footnotesize param-name}
  \textit{[}
  \texttt{[}
  \textit{\footnotesize albnd$_1$} \textit{[}\texttt{..} \textit{\footnotesize aubnd$_1$}\textit{]} \texttt{,}
  \textit{\footnotesize albnd$_2$} \textit{[}\texttt{..} \textit{\footnotesize aubnd$_2$}\textit{]}
  \dots
  \texttt{]}
  \textit{]}
  \textit{[}\texttt{=} \textit{\footnotesize values}
  \textit{]}
}

\medskip
\noindent Examples:
\begin{lstlisting}[language=stencil]
float param a
double param b[2]
float param c = 1.23
double param d[2..5, 3] = {{0.1,0.2,0.3,0.4}, {1.1,2.2,3.3,4.4}, {5,6,7,8}}
\end{lstlisting}

\textbf{param}s can be simple scalar values or, as \textbf{grid}s, (multi-dimensional) arrays of fixed size.
Furthermore, default values can be provided. If the \textbf{param} is an array, the default values are grouped
by curly braces (cf. the initialization of \texttt{d}).


\subsubsection{Iteration Space}

The iteration space, i.e., the domain of the grids on which the stencil computation is executed, is defined by
\example{
  \texttt{domainsize = (}
  \textit{\footnotesize lbnd$_1$} \texttt{..} \textit{\footnotesize ubnd$_1$} \texttt{,}
  \textit{\footnotesize lbnd$_2$} \texttt{..} \textit{\footnotesize ubnd$_2$} \dots
  \texttt{);}
}

\noindent Again, the upper bounds are inclusive. The domain size definition also defines the dimensionality of the stencil.


\subsubsection{Number of Timesteps}

The number of timesteps to be performed within one call to the generated stencil kernel function is defined by
\example{
  \texttt{iterate while} \textit{\footnotesize condition} \texttt{;}
}

\noindent Currently, the \textit{condition} must have the form ``\texttt{t <} \textit{num-timesteps}'' where
\textit{num-timesteps} is an integer literal.
The statement is optional. If it is not specified, by default, one timestep will be performed (this corresponds to
``\texttt{iterate while t < 1}'').

In the future, reduction-based stopping criteria will also be supported, which, e.g., can check for convergence.


\subsubsection{The ``operation''}

In the \textbf{operation} block, the actual stencil computation is specified which is executed on the inner
points of the grid. An operation can contain one or more statements within curly braces.
Statements can be either computations of temporary values or assignments to read-and-write grids
defined in the stencil's parameter list.
Any of arithmetic operators (\texttt{+}, \texttt{-}, $\ast$, \texttt{/}, \texttt{\textasciicircum}) can be used
(for addition, subtraction, multiplication, division, and exponentiation).

Grids are indexed as follows:
\example{
  \textit{\footnotesize grid-name} \texttt{[}
  \texttt{x} \textit{[ (} \texttt{+}$|$\texttt{-} \textit{) \textit{\footnotesize offset$_x$} ]} \texttt{,}
  \texttt{y} \textit{[ (} \texttt{+}$|$\texttt{-} \textit{) \textit{\footnotesize offset$_y$} ]} \texttt{,}
  \dots\\
  \textit{[} \texttt{;}
  \texttt{t} \textit{[ (} \texttt{+}$|$\texttt{-} \textit{) \textit{\footnotesize time-offset} ]}
  \textit{]}\\
  \textit{[} \texttt{;} \textit{\footnotesize idx$_1$} \texttt{,} \textit{\footnotesize idx$_2$} \dots  \textit{]}  
  \texttt{]}
}
\noindent i.e., by spatial coordinates first, then by the temporal index (which must not be present if the grid
was declared as \texttt{const}), and then by array indices if the grid was declared to be an array.

The offsets to the spatial identifiers (\texttt{x}, \texttt{y}, \texttt{z}, \texttt{u}, \texttt{v}, \texttt{w},
\texttt{x}\textit{i} for some integer \textit{i}) and to the spatial identifier \texttt{t} have to be compile-time constants.
The array indices also have to be compile-time constants (e.g., integer literals).

\medskip
\noindent Examples for accessing the grids defined above:
\begin{lstlisting}[language=stencil]
A[x,y+1,z-2; t+2]    // assuming that the stencil dimensionality is 3
B[x+1,y,z]           // B is const => no time component
C[x+1; t-1]
D[x,y,z; 0]          // D is const => no time component, but an array index
E[x,y,z; t+1; 0,-1]  // time component and two-dimensional array index
\end{lstlisting}

Furthermore, temporary variables (scalars or arrays) can be declared and assigned expressions.
Assigning values to arrays works as described for stencil arguments.

The stencil \textbf{operation} can also contain reductions (sums or products), which are resolved at compile time.
This feature is syntactic sugar which can save some typing.
The syntax is
\example{
  \texttt{\{}
  \textit{\footnotesize idxvar$_1$}\texttt{=}\textit{\footnotesize lbnd$_1$}\texttt{..}\textit{\footnotesize ubnd$_1$} \texttt{,}
  \textit{\footnotesize idxvar$_2$}\texttt{=}\textit{\footnotesize lbnd$_2$}\texttt{..}\textit{\footnotesize ubnd$_2$} \dots
  \textit{[} \texttt{:} \textit{\footnotesize constraints} \textit{]}
  \texttt{\}}\\
  \textit{(} \texttt{sum} $|$ \texttt{product} \textit{)}
  \texttt{(} \textit{\footnotesize expression} \texttt{)}
}
\noindent The set defines the index space for the sum or the product, and the expression within the sum or the product
the expression, possibly depending on index variables, which is summed up or multiplied for all index variable values.
Optionally, separated by a colon, in the index set definition, constraints can be specified, which are logical expressions
separated by commas.

The following example shows an \textbf{operation} which uses both a temporary array and a sum:

\begin{lstlisting}[language=stencil]
operation {
  float c[0..2] = {
    2 - 15/2 * fDT_DX_sq,
    4/3 * fDT_DX_sq,
    -1/12 * fDT_DX_sq
  };
    
  U[x,y,z; t+1] = c[0]*U[x,y,z; t] - U[x,y,z; t-1] + (
    { i=-1..1, j=-1..1, k=-1..1, r=1..2 : i^2+j^2+k^2==1 } sum(
      c[r] * U[x+r*i,y+r*j,z+r*k; t]
    )
  );
}
\end{lstlisting}
Here, a temporary array \texttt{c} is declared and initialized, and in the second part a compile time reduction is used
to sum up the neighboring points of the center point:
The three index variables \texttt{i}, \texttt{j}, \texttt{k} take the values -1, 0, and 1, but all combinations are filtered
except if \texttt{i}$^2+$\texttt{j}$^2+$\texttt{k}$^2$=1. \texttt{r} takes the values 1 and 2.
Note that the index of \texttt{c} and the spatial offsets in the grid \texttt{U} within the \texttt{sum} are compile-time
constants (the variables appear only in the sum's index set).


\subsubsection{Boundary Conditions}

The \textbf{boundaries} block can be used to specify special stencils which are to be evaluated in boundary regions.
A boundary point is characterized by having one or more spatial coordinates in the grid access fixed to discrete values.
For instance:

\begin{lstlisting}[language=stencil]
boundaries {
  U[0,y,z;t+1] = 0;
  U[x,y,z; t+1 : 0<=x<=1] = 1;
  U[x,0,z;t+1] = U[x,y_max-1,z;t];
}
\end{lstlisting}
In first statement of the example, all the points with 0 x-coordinates will be set to zero.
In the second statement all the points for which the x-coordinate is either 0 or 1 will be set to 1 (the colon notation is the
same as mentioned previously for compile-time reductions). ``\texttt{0<=x<=1}'' is a shorthand for ``\texttt{0<=x \&\& x<=1}''.
The third statement shows that the right-hand side does not have to be a constant, but can be in fact any arithmetic expression.

If the boundary block is omitted, no boundary handling is implemented in the generated stencil function.


\subsubsection{Initial Condition}

In the \textbf{initial} condition specification, the temporal coordinates of grids are fixed (as opposed to the spatial ones in the
boundary specification). All the grids read in the \texttt{operation} should be initialized with 0 substituted for \texttt{t}.
\texttt{initial} is optional; if it is not provided, \textsc{Patus} creates a default initialization with arbitrary values.

Again, as in the \textbf{boundary} block, the set-builder notation can be used to select a number of grid points to initialize.
E.g., in the last line in the example below, all the points within a cylinder around the origin with radius 30 parallel to the z-axis
are selected.
No error is raised if points are selected that are not contained in the actual grid.
The first statement shows that points can be initialized depending on their grid coordinates.

\begin{lstlisting}[language=stencil]
initial {
  U[x,y,z; -1] = sinf(2*~$\pi$~*((x-1)*fDX+fMin)) *
    sinf(2*~$\pi$~*((y-1)*fDX+fMin)) * sinf(2*~$\pi$~*((z-1)*fDX+fMin));
  U[x, y, z; 0] = U[x, y, z; -1];
  U[x, y, z; 1] = 0;

  U[x,y,z;-1 : 10<=x<=20, 5<=y<=10] = 0;
  U[x,y,z; 0 : x^2+y^2 <= 900] = 0;
}
\end{lstlisting}


\subsection{Integrating Into Your Own Application}


In the generated code (usually the file implementing the stencils in C is called \texttt{kernel.c}),
search for the function with the same name as the name of your stencil to determine the signature.
This is the function that should be called from within your application code.

See the directory \texttt{examples/applications} for examples how to integrate \textsc{Patus}-generated code
into user code.
The examples also contain Makefiles that show how the process of generating and automatically tuning stencil kernels
and use them to build the final application executable can be facilitated.
The examples are described briefly below.


%Create Fortran example