\chapter{Current Limitations}

In the current state, there are several limitations to the \textsc{Patus} framework:
\begin{itemize}
	\item Only shared memory architectures are supported (specifically: shared memory CPU systems and single-GPU setups).
	\item It is assumed that the evaluation order of the stencils within one spatial sweep is irrelevant.
		Also, always all points within the domain are traversed per sweep. One grid array is read and another
		array is written to. Such a grid traversal is called a Jacobi iteration.
		In particular, this rules out schemes with special traversal rules such as red-black Gauss-Seidel
		iterations.
	\item No extra boundary handling is applied. The stencil is applied to every interior grid point, but not to boundary
		points. I.e., the boundary values are kept constant; this corresponds to Dirichlet boundary conditions.
		To implement boundary conditions, they could be factored into the stencil operation by means of extra coefficient grids.
		In the same way, non-rectilinear domain shapes and non-uniform grids can be emulated by providing the shape and/or
		geometry information encoded as coefficients in additional grids.
		Alternatively, special $(d-1)$-dimensional stencil kernels could be specified for the boundary treatment,
		which are invoked after the $d$-dimensional stencil kernel operating on the interior. This approach, however,
		will invalidate temporal blocking schemes.
	\item The index calculation assumes that the stencil computation is carried out on a flat grid (or a grid which is
		homotopic to a flat grid). In particular, currently no spherical or torical geometries are implemented, which require
		modulo index calculations.
	\item There is no support for temporally blocked schemes yet.
%	\item Fortran limitations: only one timestep, no output pointer
%GPU limitations: only one timestep, no global sync from within kernel (global barriers cannot be programmed as blocks are
%scheduled sequentially to the multiprocessors; one block might wait being scheduled until another block completes;
%if they wait on each other the code deadlocks)
\end{itemize}

